---
title: "Practice Analysis Microbiome"
author: "Cath Burke"
date: "1 August 2017"
output: html_document
---


```{r load packages}
library("phyloseq")
library("ggplot2")
library("vegan")
library("reshape2")
library("tidyverse")
```
#Load your data into Phyloseq
```{r Loading data}
#Open script in your working directory
#Set working directory, to source file location
#Create a folder called data in this directory - store all your raw data here
#first we assign names to filesbiom table must be
biom<-"data/otu_table_final_json.biom"
#Biom table must be in json format
#Before reading the map file, remove the hash from the first line of the QIIME mapping file
map<-read.table("data/mapping_practice_depression_ps.txt", header=TRUE, row.names=1)
#Normally would just use the read_tree command, but this particular tree is a bit different to ususal.
tree<-read_tree("data/rep_set.tre")
#import data to phyloseq
practice2017<-import_biom(biom, treefilename=tree)
practice_map<-sample_data(map)
practice2017<-merge_phyloseq(practice2017,practice_map)
practice2017
colnames(tax_table(practice2017)) = c("Domain", "Phylum", "Class", "Order", "Family", "Genus", "Species")
```
#Explore the data
```{r, explore data}
#look at elements of taxonomy table and otu table
#This shows the first 10 rows of the taxonomy table for your phyloseq object
tax_table(practice2017)[1:10]
#You can subset to a specific row by name - insert a row name in the ""
tax_table(practice2017)["",]
#this returns the first 10 rows and columns of the otu table
otu_table(practice2017)[1:10,1:10]
#plot data - generic plotting options in phyloseq.
#plot_bar(practice2017)
#plot_bar(practice2017,x="Disease_status")
#plot_bar(practice2017,x="Disease_status", fill="Phylum")
#You can also use ggplot for better plotting, but takes a bit of data wrangling, more on that below.

#Five number summary of sequence coverage of samples
fivenum(colSums(otu_table(practice2017)))
#View number of sequences per sample
seq_sums<-data.frame(sort(sample_sums(practice2017)))
```
#Rarefaction analysis
```{r, rarefaction}
#Create a rarefaction curve to find the appropraite place for rarefying your data. Uses vegan for calculating the curve, and base R for plotting. 
#Plot rarecurve
t_otu_practice2017<-as.matrix(t(otu_table(practice2017)))
#col <- c("forestgreen", "red", "hotpink", "black", "blue")
#cols<-col[sinus2017_data$Type]
curve<-rarecurve(t_otu_practice2017, step=500, label=FALSE, xlim= c(0,12000), ylim=c(0,1500), ylab = "OTU Count", xaxt="n")
axis(1, xaxp=c(0, 12000, 10))
#place a line over the curve where you proppose to rarefy, in this example I used 6000 sequences
abline(v=5381, col="red")
#Then rarefy at the appropriate depth - in this example I used 6000 sequences.
practice2017_rare<-rarefy_even_depth(practice2017, sample.size=5381, rngseed=711, replace=FALSE, trimOTUs=TRUE)
```
#subsetting phyloseq objects
```{r, subsetting phyloseq objects}
#You many want to remove certain samples or OTUs from the analysis.  There are several ways to do this in Phyloseq, all require creating a new Phyloseq object.
#First example, subset to samples that match a string in a column of the sample data (i.e. the mapping file)
practice2017_rare_controls<-subset_samples(practice2017_rare, Disease_status=="control")
sample_data(practice2017_rare_controls)
#Second example, subset to samples that match everything except the string in a column of the sample data
practice2017_rare_disease<-subset_samples(practice2017_rare, Disease_status!="control")
#Now check the sample data for your new object to make sure it did what you think it did.  In this example I'm only returning the column that lists Disease_status
sample_data(practice2017_rare_disease)$Disease_status
#Third example, subset to samples that match several strings in a column of the sample data
practice2017_rare_disease2<-subset_samples(practice2017_rare, Disease_status%in%c("mdd", "recovering_mdd"))
#And check the sample data as before.
sample_data(practice2017_rare_disease2)$Disease_status
#Forth example, subset to samples that match everything except several strings in a column of the sample data
practice2017_rare_controls2<-subset_samples(practice2017_rare, !(Disease_status%in%c("mdd", "recovering_mdd")))
#And check the sample data as before.
sample_data(practice2017_rare_controls2)$Disease_status
```

#Alpha diversity analysis
```{r, alpha diversity}
#Once you have rarefied your data, explore the level of alpha diversity.  Uses phyloseq plot function which is built on ggplot2.  The x value will determine how samples are grouped along the x axis.
plot_richness(practice2017_rare, x="Disease_status", measures=c("Observed", "Chao1","Shannon"))+
  geom_boxplot()+xlab("Group")+ylab("Diversity")
#Convert alpha diversity estimates to a data form we can use for statistical testing
practice2017_rare_richness<-as.matrix(estimate_richness(practice2017_rare, measures=c("Observed","Chao1","Shannon")))
#row.names(practice2017_rare_richness)<-row.names(sample_data(practice2017_rare))
practice2017_rare_richness<-cbind(practice2017_rare_richness,sample_data(practice2017_rare))
practice2017_rare_richness
#For comaparisons of alpha diversity of more than two groups, use the Kruskal-Wallis test 
#Test if alpha diversity (observed) is significantly different between Disease_status groups
kruskal.test(Observed ~ Disease_status, data=practice2017_rare_richness)
#Now test if diversity is significantly different between controls and patients who were ever depressed (and recovering mdd groups (Original_status)
#only two groups here so use Man-Whitney U test

#wilcox.test(Observed ~ Disease_status, data=practice2017_rare_richness)
```
###UP TO HERE - DO EXAMINATION OF NEGATIVE AND POSITIVE CONTROLS
##Data cleaning and transformation
```{r data cleaning and transformation}
#Create an object with realtive abundances instead of counts
practice2017_rare_rel<-transform_sample_counts(practice2017_rare, function(x) x/sum(x))
#Filter OTUs present at less that 0.001% mean abundance
practice2017_rare_rel_fil1<-filter_taxa(practice2017_rare_rel, function(x) mean(x) > 1e-5, TRUE)
```

##Plotting functions
```{r plotting}
#I use the psmelt function to get the data into a form that ggplot likes, dplyr to filter that data I want, and ggplot to do the plotting
#Melt data into a "long" form
practice_smelt<-psmelt(practice2017_rare_rel_fil1)
head(practice_smelt)
#Let's plot the overall taxa by group (average relative abundance per group) and colour by phyla
#plot every sample by phyla, faceting into disease groups
ggplot(practice_smelt, aes(x=Sample, y=Abundance, fill=Phylum))+geom_bar(stat="identity")+facet_grid(.~Disease_status, scales="free")
#calculate the sum of abundance for each phyla per patient, then average out across groups and plot.
phyla_plot<-practice_smelt %>% group_by(Sample, Phylum, Disease_status) %>%
  summarise(phylum_sum=sum(Abundance))%>%
  group_by(Disease_status, Phylum) %>% summarise(phylum_av=mean(phylum_sum))
ggplot(phyla_plot, aes(x=Disease_status, y=phylum_av, fill=Phylum)) + geom_bar(stat="identity")+facet_grid(.~Disease_status, scales="free")+
  guides(fill=guide_legend(ncol=1))+theme(axis.text.x = element_text(angle = 90, hjust = 1),legend.key.size = unit(0.5,"line")) +ggtitle("Average Phyla per Group")



```

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
